How cryptic is cryptic diversity? Machine learning approaches to fine scale variation in the morphology of \textit{Emys marmorata}.

### 2013-04-08 documents/turtle_report.Rnw
report for meeting done

\section{Preliminary results}
\subsection{Unsupervised}
Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately \Sexpr{max(tmorph.km$clustering)} after which there is a marginal decrease in gap statistic (Fig. \ref{fig:gap}). It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important. Additionally, all gap statiic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.

@@

Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions (Table \ref{tab:dac}).

@@

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/gm}
  \caption{Visualization of PCA of \textit{E. marmorata}. The lower triangle is the pairwise comparison of the first three pricipal components. The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
  \label{fig:gm}
\end{figure}

@@

I'm currently holding back on showing the tables because there are a lot of them. I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})

@@

for(ii in seq(length(mt))) {
  caption(mt[[ii]]) <- paste0('Confusion matrix for ',
                              names(mt)[ii],
                              ' classification scheme. Rows correspond to predicted class. Columns correspond to reference class.')
  label(mt[[ii]]) <- paste0('tab:multi-conf-', names(mt)[ii])
}

for(jj in seq(length(mt))) {
  print.xtable(mt[[jj]]
               , table.placement = 'ht')
}

@

@@

\section{Future}
\begin{enumerate}
  \item remove juvelines
  \item look into multinomial logistic mixed-effects models
  \item other unsupervised methods, though that might have to wait for a follow up paper
    \begin{itemize}
      \item bayesian nonparametrics for categorical data
    \end{itemize}
\end{enumerate}

In general, misclassification seems to be at random. This is interesting for a few reasons. The assignments based on geography do not seem to bias results. All the taxa are extremely similar, though there are differences, hence the ~70\% accuracy. None of the classification schemes seems necessary better than any of the others, though sh3 is probably the best (4 classes) over all because it is the least ``random''. I think there is a summary class specific statistic that better explains this (sensitivity? specificity? detection rate?).

@@

\subsection{Grants}
\subsubsection{Hinds Fund}
Resubmit previous application with the results from this study in the prelim-results section?

Get network for one turtle? This should take long, just requires time.

\subsubsection{Paleontological Society}
Why not? Need to wait till next year. Should probably join PaleoSoc anyway.

>>>>

\section{Preliminary results}
\subsection{Unsupervised}
Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately \Sexpr{max(tmorph.km$clustering)} after which there is a marginal decrease in gap statistic (Fig. \ref{fig:gap}). It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important. Additionally, all gap statistic values are within \Sexpr{max(tmorph.gap$Tab[, 3]) - min(tmorph.gap$Tab[, 3])} of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.

@@

Dissimilarity based EAC estimated approximately \Sexpr{max(tmorph.dac)} optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference. I need to increase the number of resamples from 100 to 500 because that is when approximate stability is estimated to occur.

@@

Here is the PCA plot of the turtle pastra (Fig. \ref{fig:gm}). Also included is the mean shape and the dispersion around the landmarks. I haven't put nay classification labels on the points because an optimal partition scheme has not been selected.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/gm}
  \caption{Visualization of PCA of \textit{E. marmorata}. The lower triangle is the pairwise comparison of the first three principal components. The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
  \label{fig:gm}
\end{figure}

@@

I'm currently holding back on showing all the tables and model selection procedures because there are a lot of them. I'll show the confusion matrices from the multinomial logistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks}). The confusion matrices from the neural network and random forest models are easy to make too, just would increase the overall length of this document by quite a bit.

@@

for(ii in seq(length(mt))) {
  caption(mt[[ii]]) <- paste0('Confusion matrix for ',
                              names(mt)[ii],
                              ' classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.')
  label(mt[[ii]]) <- paste0('tab:multi-conf-', names(mt)[ii])
}

for(jj in seq(length(mt))) {
  print.xtable(mt[[jj]]
               , table.placement = 'ht')
}
@

@@

\section{Future}
\begin{enumerate}
  \item remove juveniles
  \item look into multinomial logistic mixed-effects models
  \item other unsupervised methods, though that might have to wait for a follow up paper
    \begin{itemize}
      \item bayesian nonparametrics for categorical data. Probably the best option and most interesting technically. Additionally, relatively unknown in morphological contexts though beginning to be known in biological contexts as a whole.
    \end{itemize}
\end{enumerate}

In general, misclassification seems to be at random. This is interesting for a few reasons. The assignments based on geography do not seem to bias results. All the taxa are extremely similar, though there are differences, hence the ~70\% accuracy. None of the classification schemes seems necessary better than any of the others, though sh3 is probably the best (4 classes) over all because it is the least ``random''. I think there is a summary class specific statistic that better explains this (sensitivity? specificity? detection rate?). These will illuminate which classes are ``strongest''.

@@

\subsection{Grants}
\subsubsection{Hinds Fund}
 NA terrestrial mammal community change in the Eocene. Using clustering/unsupervised methods, classify the sites in the Eocene. Spatio-temporal dynamics? Using co-occurrence networks and bipartite site-networks to understand spatio-temporal dynamics in paleo-community structure.

I have similar/preliminary results from my project from Mike Foote's class.

\subsubsection{Paleontological Society}
Why not? Need to wait till next year.

### 2013-04-08 documents/turtle_report.tex
report for meeting done

\section{Preliminary results}
\subsection{Unsupervised}
Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 40 after which there is a marginal decrease in gap statistic (Fig. \ref{fig:gap}). It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important. Additionally, all gap statiic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.

@@

Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}).

% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Thu Apr  4 16:52:02 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rr}
  \hline
 & tmorph.dac \\ 
  \hline
1 & 722 \\ 
  2 &   1 \\ 
  3 &   1 \\ 
  4 &   1 \\ 
  5 &   1 \\ 
   \hline
\end{tabular}
\caption{Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC. Each column corresponds to a different partition, with the number assigned directly below it.}
\label{tab:dac}
\end{center}
\end{table}

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/gm}
  \caption{Visualization of PCA of \textit{E. marmorata}. The lower triangle is the pairwise comparison of the first three pricipal components. The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
  \label{fig:gm}
\end{figure}

@@

I'm currently holding back on showing the tables because there are a lot of them. I'll show the confusion matrix for the multinomial logicistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks})

% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Thu Apr  4 16:52:02 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & CCR & marm & pall \\ 
  \hline
CCR &  40 &  14 &  11 \\ 
  marm &  25 &  74 &   2 \\ 
  pall &   3 &   1 &  10 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
\label{tab:multi-conf-sh1}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Thu Apr  4 16:52:02 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & CCR & marm & pall \\ 
  \hline
CCR &  44 &  12 &  10 \\ 
  marm &  17 &  82 &   6 \\ 
  pall &   2 &   1 &   7 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
\label{tab:multi-conf-sh2}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Thu Apr  4 16:52:02 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrrr}
  \hline
 & CCR & marm & pall & SJ \\ 
  \hline
CCR &  37 &  11 &   9 &   2 \\ 
  marm &  18 &  59 &   3 &  10 \\ 
  pall &   7 &   2 &  10 &   0 \\ 
  SJ &   1 &   4 &   1 &   7 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
\label{tab:multi-conf-sh3}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Thu Apr  4 16:52:02 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrrr}
  \hline
 & 1 & 2 & 3 & 4 \\ 
  \hline
1 & 107 &   9 &  13 &  13 \\ 
  2 &   1 &   8 &   0 &   0 \\ 
  3 &   4 &   0 &   8 &   3 \\ 
  4 &   4 &   0 &   2 &   8 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class.}
\label{tab:multi-conf-spinks}
\end{center}
\end{table}

>>>>

\section{Preliminary results}
\subsection{Unsupervised}
Comparison of gap statistic over a very wide range of plausible partitions indicates that as the number of partitions increase, there is a marginal increasing gap statistic until approximately 40 after which there is a marginal decrease in gap statistic (Fig. \ref{fig:gap}). It is notable that the standard errors around the gap statistic values are very large, and the marginal increases in gap statistic with an increased number of partitions may not be important. Additionally, all gap statistic values are within 0.0064 of each other meaning that there is little over all consensus for how many clusters are present when comparing gap statistics.

@@

Dissimilarity based EAC estimated approximately 5 optimal partitions (Table \ref{tab:dac}) though this may be only a marginal difference. I need to increase the number of resamples from 100 to 500 because that is when approximate stability is estimated to occur.

% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Mon Apr  8 09:29:56 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rr}
  \hline
 & tmorph.dac \\ 
  \hline
1 & 722 \\ 
  2 &   1 \\ 
  3 &   1 \\ 
  4 &   1 \\ 
  5 &   1 \\ 
   \hline
\end{tabular}
\caption{Number of specimens assigned to optimal number of partitions as determined by dissimilarity based EAC. Each column corresponds to a different partition, with the number assigned directly below it.}
\label{tab:dac}
\end{center}
\end{table}

Here is the PCA plot of the turtle pastra (Fig. \ref{fig:gm}). Also included is the mean shape and the dispersion around the landmarks. I haven't put nay classification labels on the points because an optimal partition scheme has not been selected.

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/gm}
  \caption{Visualization of PCA of \textit{E. marmorata}. The lower triangle is the pairwise comparison of the first three principal components. The upper left corner is the comparison of landmark dispersion for all specimens compared to the mean shape in red.}
  \label{fig:gm}
\end{figure}

@@

I'm currently holding back on showing all the tables and model selection procedures because there are a lot of them. I'll show the confusion matrices from the multinomial logistic regressions because they are easy (Tables \ref{tab:multi-conf-sh1}, \ref{tab:multi-conf-sh2}, \ref{tab:multi-conf-sh3}, \ref{tab:multi-conf-spinks}). The confusion matrices from the neural network and random forest models are easy to make too, just would increase the overall length of this document by quite a bit.

% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Mon Apr  8 09:29:56 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & CCR & marm & pall \\ 
  \hline
CCR &  40 &  14 &  11 \\ 
  marm &  25 &  74 &   2 \\ 
  pall &   3 &   1 &  10 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh1 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
\label{tab:multi-conf-sh1}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Mon Apr  8 09:29:56 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & CCR & marm & pall \\ 
  \hline
CCR &  44 &  12 &  10 \\ 
  marm &  17 &  82 &   6 \\ 
  pall &   2 &   1 &   7 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh2 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
\label{tab:multi-conf-sh2}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Mon Apr  8 09:29:56 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrrr}
  \hline
 & CCR & marm & pall & SJ \\ 
  \hline
CCR &  37 &  11 &   9 &   2 \\ 
  marm &  18 &  59 &   3 &  10 \\ 
  pall &   7 &   2 &  10 &   0 \\ 
  SJ &   1 &   4 &   1 &   7 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for sh3 classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
\label{tab:multi-conf-sh3}
\end{center}
\end{table}
% latex table generated in R 2.15.3 by xtable 1.7-0 package
% Mon Apr  8 09:29:56 2013
\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrrr}
  \hline
 & 1 & 2 & 3 & 4 \\ 
  \hline
1 & 107 &   9 &  13 &  13 \\ 
  2 &   1 &   8 &   0 &   0 \\ 
  3 &   4 &   0 &   8 &   3 \\ 
  4 &   4 &   0 &   2 &   8 \\ 
   \hline
\end{tabular}
\caption{Confusion matrix for spinks classification scheme. Rows correspond to predicted class. Columns correspond to reference class. From multinomial logistic regression models.}
\label{tab:multi-conf-spinks}
\end{center}
\end{table}
